Elastic Scattering in General Relativistic Ray Tracing for Neutrinos

### 2017-08-01 paper.tex
Make consistent your use of \tilde for Eulerian quantities

A stationary observer sees a neutrino with momentum
$p_\alpha \rightarrow \tilde{\varepsilon}(-1,\Omega_i)$,
with direction $\Omega_i \rightarrow (\sin A, 0, \cos A)$,
and observed energy $\tilde{\varepsilon}$.
In the fluid frame this neutrino has energy $\varepsilon=-u^\alpha p_\alpha$;
therefore the average energy varies with observing angle:
\begin{equation}
  \label{eqn:doppler_avg_eps}
  \langle \tilde{\varepsilon} \rangle(\cos A) =
  \langle \varepsilon \rangle^{\rm eq} \frac{1}{W(1-v \cos A)}.
\end{equation}
The equilibrium average energy $\langle \varepsilon \rangle^{\rm eq}$
is given by $T \mathscr{F}_3(\eta_\nu)/\mathscr{F}_2(\eta_\nu)\approx 3T$,
with the Fermi integrals given in Eqn.~\ref{eqn:fermi_integral}.
Eqn.~\ref{eqn:doppler_avg_eps} describes the well-known Doppler effect.

@@

Since no analytic form is known for the background field $\Phi$,
we examine only the \emph{noscat} case,
with $\chi_s=0$ and $\chi=\chi_a^{*}$.
This scenario has been widely used in the literature as a test for
radiation codes
\cite{smit1997-two_moment,abdi2012-monte_carlo,fouc2015-m1_nsbh}.
We sample the distribution function over a uniform grid of
30 points in angle $\cos A \in [0.734,1]$ and
40 points in energy $\varepsilon \in [0,100]$~MeV.
Because of the step function in fluid variables at radius $R$,
we throttle the maximum step size at 1.25~km.

@@

Though we do not display it here,
we also examine $\langle \varepsilon \rangle(\cos A)$
as in Eqn.~\ref{eqn:avg_eps_per_cosA}, and robustly reproduce the
common feature of increasing average energies at the limb of the star
for $\nu_e$ and $\bar{\nu}_e$. In contrast $\langle \varepsilon \rangle$
for $\nu_x$ is constant across angles, as expected when the star
is optically thin at even the highest energies $\varepsilon\sim100$~MeV.

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Optically thick star in Schwarszchild.
Total luminosity at infinite separation may be computed by
computing the radial energy flux at a finite separation from the star, $r$,
and using $p_t$ for the energy. We compute this asymptotic luminosity
at several radii, $r$, and show that the luminosity is independent of
the observer's radius.
\footnote{TODO: compute $L$, put in figure}

@@

Total luminosities and average energies from the different treatments are
\footnote{TODO: update MC results 02->03}
presented in Tab.~\ref{tab:collapse_rt_vs_mc}.
As expected, $\nu_e$ is least affected by scattering, and $\nu_x$ most.
In fact, without scattering, $\nu_x$ luminosities are overestimated by more than
two orders of magnitude, due to the steep increase in temperature with depth
in the inner core.
Though contributing only a fraction of the total luminosity,
the average energy of neutrinos scattered to the observer from the envelope
outside the shock, $\langle\varepsilon\rangle^{\rm sc}$,
is poorly characterized by the \emph{noscat} treatment for all species.
By contrast, both the \emph{gray} and \emph{spectral} \emph{scat} treatments
faithfully describe the high scattered energies from the envelope.

\begin{table}%[h]
  % to get mc luminosities: tail $MCDATA/Spectra/luminosities_and_avg_eps-501km-nu?.dat
  % to get rt luminosities: tail $RTDATA/Moments-nu?.out
  % MCDATA=$DIR/ResultsSherwood-02
  % RTDATA_NO=$DIR/ResultsBrett-B-03/Test-noscat-01/
  % RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
  % RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
  % DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20161007-ray_tracing_tests/CollapseProfile/
  \caption{
    Comparison of total luminosities and average energies between
    the following methods:
    ray tracing ignoring elastic scattering `RT noscat',
    ray tracing with the gray scattering treatment `RT scat gr',
    ray tracing with the spectral scattering treatment `RT scat sp', and
    a fiducial Monte Carlo transport evolution `MC'.
    We also show average energies of the scattering envelope
    $\langle\varepsilon\rangle^{\rm sc}$, which are estimated by eye
    from plots like Fig.~\ref{fig:avg_eps_collapse} at $\cos A < 0.95$.
    Luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$ and
    average energies have units MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:collapse_rt_vs_mc}
  \begin{tabularx}{\columnwidth}{X r r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat gr} & \,\,{\bf RT scat sp} & \,\,{\bf MC} \\
    \hline
    $L_{\nu_e}$                                         &  3.53 & 3.14 & 3.04 & 3.48 \\
    $L_{\bar{\nu}_e}$                                   & 5.19 & 3.05 & 3.03 & 3.01 \\
    $L_{\nu_x}$                                         & 222.0  & 1.88 & 1.76 & 1.70 \\
    $\langle \varepsilon_{\nu_e} \rangle$               &  10.6 & 10.6 & 10.9 & 11.0 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle$         & 14.2 & 13.0 & 13.8 & 13.7 \\
    $\langle \varepsilon_{\nu_x} \rangle$               & 47.8 & 16.0 & 17.3 & 16.2 \\
    $\langle \varepsilon_{\nu_e} \rangle^{\rm sc}$      & 4    & 14   & 17   & 16 \\
    $\langle \varepsilon_{\bar{\nu}_e} \rangle^{\rm sc}$& 4    & 24   & 20   & 18 \\
    $\langle \varepsilon_{\nu_x} \rangle^{\rm sc}$      & 2    & 28   & 31   & 27 \\
    \hline
  \end{tabularx}
\end{table}

@@

By expanding the exponential of
Eqns.~\ref{eqn:fse_homogeneous_general} and \ref{eqn:fae_homogeneous_general}
as we did for Eqn.~\ref{eqn:fae_homogeneous_limits},
and furthermore factoring out the dominant energy dependence from the
cross-sections (i.e.\ $\chi\equiv\varepsilon^2 \zeta$
with $\zeta$ approximately constant), we have
\begin{align}
  \label{eqn:fae_homogeneous_thin_limit}
  f_{\rm AE}(\varepsilon)
  &= \varepsilon^2 \zeta_a \,f^{\rm eq}(\varepsilon)\,s, \\
  \label{eqn:fse_homogeneous_thin_limit}
  f_{\rm SE}(\varepsilon)
  &= \varepsilon^2 \zeta_s \,\Phi(\varepsilon)\,s,
\end{align}
where $s$ is the length of scattering envelope passed through by the ray.
\footnote{TODO: should be $\chi_a^*=\varepsilon\zeta_a$ for $\nu_x$}
In the envelope the local temperature is low so $f^{\rm eq}/\Phi\ll 1$,
and also $\zeta_a/\zeta_s\ll 1$,
so $f_{\rm SE}$ strongly dominates over $f_{\rm AE}$.
The average energy of the scattered field is therefore
\begin{equation}
  \label{eqn:avg_eps_sc}
  \langle \varepsilon \rangle^{\rm sc}
  = \frac{\int d\varepsilon\,\varepsilon^3 f}{\int d\varepsilon\,\varepsilon^2 f}
  \approx \frac{\int d\varepsilon\,\varepsilon^5 \Phi}{\int d\varepsilon\,\varepsilon^4 \Phi}.
\end{equation}

@@

In order to estimate $\langle \varepsilon \rangle^{\rm sc}$ analytically,
we must write the background field $\Phi(\varepsilon)$ analytically.
Direct Fermi-Dirac fits using a temperature and chemical potential
representative of a physical neutrinosurface fair poorly since neutrinos of
different energies decouple from the matter at different radii, over which
the thermodynamic state varies substantially.
Phenomenological Fermi-Dirac fits work well;
but so do pinched spectral fits which are much simpler
\cite{keil2003-pinched_spectra,miri2016-sn_neutrinos}:
\begin{equation}
  \label{eqn:pinched_spectra}
  \Phi^{\rm pi}(\varepsilon) \propto
  \varepsilon^{\alpha-2}
  \exp\left(-(\alpha+1)\frac{\varepsilon}{\langle\varepsilon\rangle}\right),
\end{equation}
(where our definition differs from that of \cite{keil2003-pinched_spectra}
by a factor of
$\varepsilon^2$, since we define our distribution function to be dimensionless
in natural units $\{\hbar,c\}=1$).

Pinching parameters
(calculated by eye from the \emph{spectral} \emph{scat} method)
for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\varepsilon\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to approximately 10\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of all of the treatments including elastic scattering presented
in Tab.~\ref{tab:collapse_rt_vs_mc}, except for
$\bar{\nu}_e$ in the \emph{gray} \emph{scat} treatment,
which deviates from our prediction by 30\%.
This agreement is excellent despite the drastic
simplifications used in our toy model.

The $\langle\varepsilon_{\bar{\nu}_e}\rangle^{\rm sc}$ prediction
in the \emph{gray} \emph{scat} treatment is approximately 30\%
larger than the the fiducial Monte Carlo estimate. This is
due to the large negative local chemical potential
$\eta_{\bar{\nu}_e}\sim-10$ in the scattering envelope.
As described in Sec.~\ref{ssec:sources_se},
we use $\eta_\nu$ in the \emph{gray} treatment to construct our
synthetic spectrum from total neutrino densities $J$ and $G$.
This error and our successful analysis using pinched spectra above,
points the way to future improvements to the \emph{gray} method:
making better assumptions about the spectra which don't depend upon
local fluid quantities.

>>>>

A stationary observer measures an energy of $\tilde{\varepsilon}$
for a neutrino with momentum
$p_\alpha \rightarrow \tilde{\varepsilon}(-1,\Omega_i)$
and direction $\Omega_i \rightarrow (\sin A, 0, \cos A)$.
In the fluid frame this neutrino has energy $\varepsilon=-u^\alpha p_\alpha$;
therefore the average energy varies with observing angle:
\begin{equation}
  \label{eqn:doppler_avg_eps}
  \langle \tilde{\varepsilon} \rangle(\cos A) =
  \langle \varepsilon \rangle^{\rm eq} \frac{1}{W(1-v \cos A)}.
\end{equation}
The equilibrium average energy $\langle \varepsilon \rangle^{\rm eq}$
is given by $T \mathscr{F}_3(\eta_\nu)/\mathscr{F}_2(\eta_\nu)\approx 3T$,
with the Fermi integrals given in Eqn.~\ref{eqn:fermi_integral}.
Eqn.~\ref{eqn:doppler_avg_eps} describes the well-known Doppler effect.

@@

Since no analytic form is known for the background field $\Phi$
interior to the star, we examine only the \emph{noscat} case,
with $\chi_s=0$ and $\chi=\chi_a^{*}$.
This scenario has been widely used in the literature as a test for
radiation codes
\cite{smit1997-two_moment,abdi2012-monte_carlo,fouc2015-m1_nsbh}.
We sample the distribution function over a uniform grid of
30 points in angle $\cos A \in [0.734,1]$ and
40 points in energy $\tilde{\varepsilon} \in [0,100]$~MeV.
Because of the step function in fluid variables at radius $R$,
we throttle the maximum step size at 1.25~km.

@@

Though we do not display it here,
we also examine $\langle \tilde{\varepsilon} \rangle(\cos A)$
as in Eqn.~\ref{eqn:avg_eps_per_cosA}, and robustly reproduce the
common feature of increasing average energies at the limb of the star
for $\nu_e$ and $\bar{\nu}_e$. In contrast $\langle \tilde{\varepsilon} \rangle$
for $\nu_x$ is constant across angles, as expected when the star
is optically thin at even the highest energies $\tilde{\varepsilon}\sim100$~MeV.

\subsection{Idealized Compact Star:
  Testing Gravitational Redshift and Geodesic Curvature}
\label{ssec:test_gravity}
Unfinished.
Optically thick star in Schwarszchild.
Total luminosity at infinite separation may be computed by
computing the radial energy flux at a finite separation from the star, $r$,
and using $p_t$ for the energy. We compute this asymptotic luminosity
at several radii, $r$, and show that the luminosity is independent of
the observer's radius.
\footnote{TODO: compute $L$, put in figure}
Also compare a few geodesics in Kerr to analytic.

@@

Total luminosities and average energies from the different treatments are
\footnote{TODO: update MC results 02->03}
presented in Tab.~\ref{tab:collapse_rt_vs_mc}.
\footnote{TODO: give equation for L}
As expected, $\nu_e$ is least affected by scattering, and $\nu_x$ most.
In fact, without scattering, $\nu_x$ luminosities are overestimated by more than
two orders of magnitude, due to the steep increase in temperature with depth
in the inner core.
Though contributing only a fraction of the total luminosity,
the average energy of neutrinos scattered to the observer from the envelope
outside the shock, $\langle\varepsilon\rangle^{\rm sc}$,
is poorly characterized by the \emph{noscat} treatment for all species.
By contrast, both the \emph{gray} and \emph{spectral} \emph{scat} treatments
faithfully describe the high scattered energies from the envelope.

\begin{table}%[h]
  % to get mc luminosities: tail $MCDATA/Spectra/luminosities_and_avg_eps-501km-nu?.dat
  % to get rt luminosities: tail $RTDATA/Moments-nu?.out
  % MCDATA=$DIR/ResultsSherwood-02
  % RTDATA_NO=$DIR/ResultsBrett-B-03/Test-noscat-01/
  % RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
  % RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
  % DIR=/Users/mbdeaton/Documents/ResearchGraveyard/NuclearAstrophysics/DataAndVis/20161007-ray_tracing_tests/CollapseProfile/
  \caption{
    Comparison of total luminosities and average energies between
    the following methods:
    ray tracing ignoring elastic scattering `RT noscat',
    ray tracing with the gray scattering treatment `RT scat gr',
    ray tracing with the spectral scattering treatment `RT scat sp', and
    a fiducial Monte Carlo transport evolution `MC'.
    We also show average energies of the scattering envelope
    $\langle\tilde{\varepsilon}\rangle^{\rm sc}$, which are estimated by eye
    from plots like Fig.~\ref{fig:avg_eps_collapse} at $\cos A < 0.95$.
    Luminosities have units $10^{52}\,{\rm erg}\,{\rm s}^{-1}$ and
    average energies have units MeV.
    The $\nu_x$ luminosities are per-species:
    multiply by four to get the total heavy-lepton neutrino luminosities.
  }
  \label{tab:collapse_rt_vs_mc}
  \begin{tabularx}{\columnwidth}{X r r r r}
    & {\bf RT noscat} & \,\,{\bf RT scat gr} & \,\,{\bf RT scat sp} & \,\,{\bf MC} \\
    \hline
    $L_{\nu_e}$                                         &  3.53 & 3.14 & 3.04 & 3.48 \\
    $L_{\bar{\nu}_e}$                                   & 5.19 & 3.05 & 3.03 & 3.01 \\
    $L_{\nu_x}$                                         & 222.0  & 1.88 & 1.76 & 1.70 \\
    $\langle \tilde{\varepsilon}_{\nu_e} \rangle$               &  10.6 & 10.6 & 10.9 & 11.0 \\
    $\langle \tilde{\varepsilon}_{\bar{\nu}_e} \rangle$         & 14.2 & 13.0 & 13.8 & 13.7 \\
    $\langle \tilde{\varepsilon}_{\nu_x} \rangle$               & 47.8 & 16.0 & 17.3 & 16.2 \\
    $\langle \tilde{\varepsilon}_{\nu_e} \rangle^{\rm sc}$      & 4    & 14   & 17   & 16 \\
    $\langle \tilde{\varepsilon}_{\bar{\nu}_e} \rangle^{\rm sc}$& 4    & 24   & 20   & 18 \\
    $\langle \tilde{\varepsilon}_{\nu_x} \rangle^{\rm sc}$      & 2    & 28   & 31   & 27 \\
    \hline
  \end{tabularx}
\end{table}

@@

By expanding the exponential of
Eqns.~\ref{eqn:fse_homogeneous_general} and \ref{eqn:fae_homogeneous_general}
as we did for Eqn.~\ref{eqn:fae_homogeneous_limits},
and furthermore factoring out the dominant energy dependence from the
cross-sections (i.e.\ $\chi\equiv\varepsilon^2 \zeta$
with $\zeta$ approximately constant), we have
\begin{align}
  \label{eqn:fae_homogeneous_thin_limit}
  f_{\rm AE}(\varepsilon)
  &= \varepsilon^2 \zeta_a \,f^{\rm eq}(\varepsilon)\,s, \\
  \label{eqn:fse_homogeneous_thin_limit}
  f_{\rm SE}(\varepsilon)
  &= \varepsilon^2 \zeta_s \,\Phi(\varepsilon)\,s,
\end{align}
where $s$ is the length of scattering envelope passed through by the ray.
\footnote{TODO: should be $\chi_a^*=\varepsilon\zeta_a$ for $\nu_x$}
In the envelope the local temperature is low so $f^{\rm eq}/\Phi\ll 1$,
and also $\zeta_a/\zeta_s\ll 1$,
so $f_{\rm SE}$ strongly dominates over $f_{\rm AE}$.
The average energy of the scattered field measured by our observer
is therefore
\begin{equation}
  \label{eqn:avg_eps_sc}
  \langle \tilde{\varepsilon} \rangle^{\rm sc}
  = \frac{\int d\varepsilon\,\varepsilon^3 f}{\int d\varepsilon\,\varepsilon^2 f}
  \approx \frac{\int d\varepsilon\,\varepsilon^5 \Phi}{\int d\varepsilon\,\varepsilon^4 \Phi}.
\end{equation}
Note that we have taken the liberty here of identifying the fluid-frame energy
$\varepsilon$ with the Eulerian energy $\tilde{\varepsilon}$,
since infall velocities in the envelope are $\sim0.1c$,
and as Fig.~\ref{fig:avg_eps_doppler} indicates,
the Doppler shift will therefore introduce an error into our analysis of
$\sim10\%$.

@@

In order to estimate $\langle \tilde{\varepsilon} \rangle^{\rm sc}$ analytically,
we must write the background field $\Phi(\varepsilon)$ analytically.
Direct Fermi-Dirac fits using a temperature and chemical potential
representative of a physical neutrinosurface fair poorly since neutrinos of
different energies decouple from the matter at different radii, over which
the thermodynamic state varies substantially.
Phenomenological Fermi-Dirac fits work well;
but so do pinched spectral fits which are much simpler
\cite{keil2003-pinched_spectra,miri2016-sn_neutrinos}:
\begin{equation}
  \label{eqn:pinched_spectra}
  \Phi^{\rm pi}(\varepsilon) \propto
  \varepsilon^{\alpha-2}
  \exp\left(-(\alpha+1)\frac{\varepsilon}{\langle\varepsilon\rangle}\right),
\end{equation}
(where our definition differs from that of \cite{keil2003-pinched_spectra}
by a factor of
$\varepsilon^2$, since we define our distribution function to be dimensionless
in natural units $\{\hbar,c\}=1$).

Pinching parameters
(calculated by eye from the \emph{spectral} \emph{scat} method)
for the species
$\{\nu_e,\bar{\nu}_e,\nu_x\}$ are $\alpha\approx\{3.6,5.1,2.3\}$.
Energy moments of pinched spectra like those in Eqn.~\ref{eqn:avg_eps_sc}
have a simple analytic form so that Eqn.~\ref{eqn:avg_eps_sc} becomes
\begin{align}
  \label{eqn:pinched_avg_eps_sc_1}
  \langle\tilde{\varepsilon}\rangle^{\rm sc}
  &\approx \frac{\alpha+3}{\alpha+1} \langle\varepsilon\rangle, \\
  %\label{eqn:pinched_avg_eps_sc_2}
  %&= \{1.4,1.3,1.6\} \langle\varepsilon\rangle, \\
  \label{eqn:pinched_avg_eps_sc_3}
  &\approx \{16,18,28\}\,~{\rm MeV},
\end{align}
again for $\{\nu_e,\bar{\nu}_e,\nu_x\}$ respectively.
These analytic predictions agree with the average energies of the scattering
envelope to approximately 10\%
% From by-eye examinations of <eps>(cosA) plot using data from
% MCDATA=$DIR/ResultsSherwood-02
% RTDATA_GR=$DIR/ResultsBrett-B-03/Test-scat-01/
% RTDATA_SP=$DIR/ResultsBrett-A-03/Test-scat-01/
%          nue     nua     nux
of all of the treatments including elastic scattering presented
in Tab.~\ref{tab:collapse_rt_vs_mc}, except for
$\bar{\nu}_e$ in the \emph{gray} \emph{scat} treatment,
which deviates from our prediction by 30\%.
This agreement is excellent despite the drastic
simplifications used in our toy model.

The $\langle\tilde{\varepsilon}_{\bar{\nu}_e}\rangle^{\rm sc}$ prediction
in the \emph{gray} \emph{scat} treatment is approximately 30\%
larger than the the fiducial Monte Carlo estimate. This is
due to the large negative local chemical potential
$\eta_{\bar{\nu}_e}\sim-10$ in the scattering envelope.
As described in Sec.~\ref{ssec:sources_se},
we use $\eta_\nu$ in the \emph{gray} treatment to construct our
synthetic spectrum from total neutrino densities $J$ and $G$.
This error and our successful analysis using pinched spectra above,
points the way to future improvements to the \emph{gray} method:
making better assumptions about the spectra which don't depend upon
local fluid quantities.
